---
title: "Congressional Stagnation - Does it Come Down to Money?"
subtitle: "SDS 192: MP2"
author: 
- name: Katelyn Diaz
  affiliation: Smith College
- name: Helen Glover
  affiliation: Smith College
- name: Misa Tetsuka
  affiliation: Smith College
date: "`r format(Sys.Date(), '%B %e, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    fig_width: 7
    fig_height: 6
    fig_caption: true
    theme: lumen
    df_print: paged
    code_folding: hide
    #was originally show
references:
- id: fec16
  type: article
  title: "fec16: Data Package for the 2016 United States Federal Elections"
  author: 
  - family: Tapal
    given: Marium
  - family: Gahwagy
    given: Rana
  - family: Ryan
    given: Irene
  - family: Baumer
    given: Benjamin S.
  issued:
    year: 2020
  publisher: CRAN
  url: https://CRAN.R-project.org/package=fec16
  notes: "R package version 0.1.3"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  echo = TRUE,
  warning = FALSE
)
library(tidyverse)
library(sds192)
library(fec16)
library(dplyr)
library(scales)
library(plotly)
library(ggthemes)
```
***
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">
Recently, politics has been at the forefront of everyone's minds. Inspired by the recent presidential election and Senate runoffs, our group wanted to look more closely at the U.S. election process. 
</div>

***

## What Is Congressional Stagnation?
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">
Generally, it is incredibly difficult for challengers to unseat incumbents. Reelection rates are high, and challengers often do not win. According to Center of Representative Politics research, the House reelection rate every four years from 1984 to 2016 was above 85%. [^1]
<br>
This high reelection rate for incumbents is called **Congressional Stagnation**, suggesting that incumbents enjoy an advantage when running for reelection. Academics suggest there is an important financial factor that plays into this advantage. Based on factors such as increased name recognition, incumbents receive more national media attention, improving their fundraising and campaign bases.[^2] Consequently, this leads to a higher presumption of success which could also deter voter turnout, as the public is led to believe that only the incumbent can win.
<br>
In our data analysis, the goal is to analyze the correlation between campaign finances and whether or not the candidate is an incumbent.
</div>
***
## 2016 US House of Representatives Election Data
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">
The `fec16` data package includes data from the Federal Election Commission's 2015-2016 election cycle. For our analysis we used the `results_house` (US House of Representatives Election Results) and `campaigns` (Campaign) datasets from this package. 
</div>
```{r, message = FALSE}
# Renaming dataset
winner <- results_house
# Clean up data and get rid of columns in winner table that we aren't using
small_data <- select(winner, -c(primary_votes, primary_percent, general_votes, general_percent, runoff_votes, runoff_percent, footnotes, district_id)) 
# want total finances that candidates received
# * ignored all transfers, disregarded total receipts bc we don't know what's in it
finances <- campaigns %>%
  mutate(campaigns,
         finances_total = (pol_pty_contrib + other_pol_cmte_contrib + ttl_indiv_contrib + cand_contrib))
# cleaning up finances data frame
final_finances <- select(finances, -c(pty_cd, ttl_receipts, trans_from_auth, ttl_disb, trans_to_auth, coh_bop, coh_cop, cand_contrib, cand_loans, other_loans, cand_loan_repay, other_loan_repay, debts_owed_by, ttl_indiv_contrib, cand_office_st, cand_office_district, other_pol_cmte_contrib, pol_pty_contrib, cvg_end_dt, indiv_refunds, cmte_refunds))
# joining the two tables
sorted_finances <- final_finances %>%
  inner_join(small_data, by = c("cand_id" = "cand_id"))
```
***
## Do Incumbent US House Candidates Receive More Campaign Contributions than Non-Incumbents? 
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">
Narrowing the focus on the two major political parties, Democrat and Republican, we examine whether incumbent candidates received a larger total amount of campaign contributions* than non-incumbent candidates.
<br>
* *Total campaign contributions is calculated based on contributions from party committees, other political committees, total individual contribution, and candidate contributions listed in the `campaigns` dataset*
</div>
```{r, message = FALSE}
finance_plot <- ggplot(sorted_finances, aes(x= cand_name, y = finances_total)) +
  geom_point(aes(color = incumbent), size = 1.5) +
  scale_color_manual(breaks = c("TRUE", "FALSE"),values=c("seagreen3","plum3"))+
  guides(color=guide_legend(title= "Incumbent"))+
  theme_stata()+
  scale_y_continuous(labels = dollar)+
   labs(title = "US House Candidate VS Total Campaign Contributions\n by Incumbency",
    x = "Candidates",
    y = "Total Campaign Contribution ($)"
  )+
  theme( axis.text.x = element_blank(), axis.ticks.x = element_blank())

ggplotly(finance_plot) 

```

<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">

The graph demonstrates the relationship between a House candidate’s total campaign contributions, and whether they are an incumbent. Each point represents an individual candidate. The y-axis displays the candidate’s total campaign contributions in US dollars. 
<br>
Based on this graph, we conclude that there is strong evidence of congressional stagnation, as the data for non-incumbents is concentrated along the bottom. While there is a similarity in contribution amounts as both candidate types have similar scattering, overwhelmingly the incumbents are seen to have raised more money than non-incumbents. There are exceptions, as in the case of David Trone who raised more than $13 million dollars ($9.1 million of his personal fortune) in his unsuccessful Democractic primary campaign (high amount was the result of trying to catch up after entering the race late). [^3]
</div>

***
## A Deeper Look Into US House Election Winners
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">

Delving deeper, we created another graph that specifically looks at only the **winners** of the 2016 US House Elections. Total campaign contributions remains to be calculated based on contributions from party committees, other political committees, total individual contribution, and candidate contributions listed in the `campaigns` dataset
</div>

```{r, message =FALSE}
#Excluding all non-winners
winners <- results_house[which(results_house$won == "TRUE"),] 
smalls_data <- select(winners, -c(primary_votes, primary_percent, general_votes, general_percent, runoff_votes, runoff_percent, footnotes, district_id)) 
# joining the two tables
sortedwon_finances <- final_finances %>%
  inner_join(smalls_data, by = c("cand_id" = "cand_id"))

finances_plot <- ggplot(sortedwon_finances, aes(x= cand_name, y = finances_total)) +
  geom_point(aes(color = incumbent), size = 1.5) +
  scale_color_manual(breaks = c("TRUE", "FALSE"), values=c("seagreen3","plum3")) +
  guides(color=guide_legend(title= "Incumbent"))+
  scale_y_continuous(labels = dollar)+
   theme_stata()+
   labs(
    title = "US House Election Winners VS Total Campaign Contributions\n by Incumbency",
    x = "Candidates",
    y = "Total Campaign Contributions ($)"
  ) +
  theme( axis.text.x = element_blank(), axis.ticks.x = element_blank())

ggplotly(finances_plot)
```
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">
Looking only at candidates that won, there is a strong sign of congressional stagnation; there are far more incumbents winners than non incumbents. Interestingly, both the incumbents and non-incumbents are similarly scattered when it comes to total amount of campaign contributions with a high concentration towards the bottom.
</div>

***
## A Deeper Look Into Party Lines in Michigan
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">
Based on the graphs above, it is clear that in general, incumbents received more campaign contributions. Diving deeper, we wanted to see if this trend is consistent across the two major party lines. We chose Michigan, a key battleground state, as we saw the state flip in the 2020 Presidential Election. No distinct correlation could be seen, which suggests that the incumbent advantage is enjoyed similarly in both parties.
<br>
*Total campaign contributions is calculated based on contributions from party committees, other political committees, total individual contribution, and candidate contributions listed in the `campaigns` dataset*
</div>

``` {r, message = FALSE}
#Excluding all non-Michigan candidates
winner_MI <- results_house[which(results_house$state == "MI"),]
#Excluding all third party candidates
winner_MID <- winner_MI[which(winner_MI$party == "DEM"),]
winner_MIR <- winner_MI[which(winner_MI$party == "REP"),]
RD_MI <- rbind(winner_MID, winner_MIR)
# Clean up data and get rid of columns in winner table that we aren't using
MI_smalldata <- select(RD_MI, -c(primary_votes, primary_percent, general_votes, general_percent, runoff_votes, runoff_percent, footnotes, district_id)) 
# joining the two Michigan tables
MI_sortedfinances <- final_finances %>%
  inner_join(MI_smalldata, by = c("cand_id" = "cand_id"))
#Plot Code
MI_finplot <- ggplot(MI_sortedfinances, aes(x= finances_total , y = cand_name)) +
  geom_point(aes(color = party, shape = incumbent), size = 3) +
  scale_color_manual(breaks = c("DEM", "REP"),
                     values=c("blue", "red")) +
  theme(plot.background = element_rect(fill = "slategray2")) +
  labs(
    title = "MI House Candidates VS Total Campaign Contributions", 
    y = "Candidates",
    x = "Total Campaign Contributions ($)"
  ) +
  scale_x_continuous(labels = dollar)
MI_finplot
```


***
## Analysis
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">
Our findings suggest that there is a correlation between a candidate's incumbency and higher amounts of total campaign contributions. The US House Candidates Plot demonstrates that the majority of non-incumbents are concentrated towards the bottom, suggesting they receive a relatively small total amount of campaign contributions. The candidates that received more than $5 million in total campaign contributions are overwhelmingly incumbents. As predicted, the incumbent advantage appears to have a significant financial factor.
<br>
	The US House Election Winners Plot further displays congressional stagnation, as the majority of winners are incumbents. Diverging from the initial graph, both the incumbents and non-incumbents are similarly scattered with a high concentration towards the bottom. For the incumbents, this could be a result of not needing to rely as heavily on contributions, as they have name recognition. For the non-incumbents, we did not filter out the open seats. As a result, non-incumbents had a good chance of winning with few contributions in such districts. For example, Valez Demings, a non-incumbent, ran and won an open seat despite only raising 160,841 dollars. Political party affiliation also plays a crucial role in helping non-incumbents; Ms. Demings, a Democratic, ran in a Democratic district.[^5]
</div>

***

## Conclusion

</style>
<div class = "blue">
Analyzing congressional stagnation by comparing the incumbency of House candidates to their total campaign contributions, our findings suggests there was an incumbent advantage in the 2016 House of Representatives election. In conjunction to receiving more campaign contributions, incumbents had a higher proportion of reelection. Further examination into party affiliation found no clear correlation, suggesting that both major political parties similarly enjoyed an incumbent advantage in the 2016 election cycle.  
</div>
***

## Word Count
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">

```{r word_count, message=FALSE, echo=FALSE}
text_stats()
```
</div>
***
## Standards
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">

In this assignment, we attempted the following [standards](https://beanumber.github.io/sds192/standards.html):

##### Necessary 

- `r standard_button("github")`: We mastered the Github standard because all members were able to clone the repository, and successfully pull, commit, and push. We were also able to resolve merge conflicts and mark and resolve issues.  

- `r standard_button("wrangling")`: We mastered the Wrangling standard because we used the verbs to slice and dice the dataset tables. By working together to read and organize the data, we then use these functions multiple times to create the desired data summaries. We also included helpful comments in our code for increased readability.


- `r standard_button("relational")`: We mastered the relational standard because we used multiple data frames, to create a new dataset relevant to our project topic by successfully joining candidates and their finances together. We also ensured that we were only joining data that we wanted, and discarding data that was not necessary. 

##### Optional

<<<<<<< HEAD
- `r standard_button("aesthetics")`: We mastered the aesthetics standards by utilizing a wide range of visual cues and colors to help strengthen the graphs’ visual appearances. For example, we have categorical color scheme colors, clear x and y axes, and background colors. By also creating interactive graphs, viewers can look at individual data points. This prevents excessive cluttering from text labeling. We also chose the colors we used purposefully. Red and blue has an intrinsic meaning and we ensured that we stuck to it when we were graphing that data, and used other colors of the same depth, when we were not. 
=======
- `r standard_button("aesthetics")`: We mastered the aesthetics standards by utilizing a wide range of visual cues and colors such as categorical color scheme colors, clear x and y axes, and background colors. By making the graph interactive, individual data points are visible, preventing excessive cluttering from labeling. We chose  colors purposefully as red and blue have an intrinsic meaning, ensuring that we used other colors when we were not discussing party affiliation. 
>>>>>>> aac7eb4eded9423e1c295189eb8997a53070f9b5

- `r standard_button("markdown")`: We mastered the R Markdown standard by using a variety of elements including two graphs to add context, external research, bold headings, and an easy to follow text format. By hiding all unnecessary and irrelevant context, our markdown has a clean appearance. We automatically hide the code while allowing our code to be easily accessible if wanted

</div>

## References

[^1]: **NPR** *Why it’s so good to be an incumbent*
(https://www.npr.org/sections/itsallpolitics/2012/06/11/154745966/why-its-good-to-be-the-incumbent)

[^2]: **Center for Representative Politics** *US House Reelection Rates, 1964 - 2018*
(https://www.opensecrets.org/elections-overview/reelection-rates )

[^3]: **Washington Post**  *David Trone spends 9.1 million of his own money* (https://www.washingtonpost.com/local/md-politics/david-trone-spends-91-million-of-his-own-money-on-md-congressional-race/2016/04/12/29f5827c-00c2-11e6-b823-707c79ce3504_story.html)

[^4]: **NPR** *This candidate is self-funding more than anymore in Congress* (https://www.npr.org/2016/04/23/475268587/this-candidate-is-spending-more-than-anyone-ever-for-a-seat-in-congress)

[^5]: **Ballotpedia** *Val Demings*
(https://ballotpedia.org/Val_Demings)




 